The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-31 manuscript/rebuttal.Rmd

RESPONSE: We appreciate reviewer 2's positive feedback, insights and thoughtful suggestions.

### R2 Point 1

> 1. Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case
> studies or practical examples that demonstrate the effectiveness of your recommendations.
> Highlighting specific instances where these best practices have enhanced reproducibility or solved
> particular issues could substantiate your points more effectively.

RESPONSE: To demonstrate the importance of computational reproducibility, we elaborate on the
"forensic bioinformatics" case study of Baggerly and Coombes (2010) @Baggerly2010-nr critique of Potti
et al (2006) @Potti2006-yk in the introduction:

@@

RESPONSE: It is possible to achieve accurate journal-specific formatting of references with a Rmarkdown
document.
It requires a few things:

* A .bib file that contains the bibliography in bibtex format.

* A .csl file that specifies the citation style.
These are avaialble from [GitHub](https://github.com/citation-style-language/styles).

* Modification to the Rmd header section that specifies the path of the bib and csl files.

@@

This enables convenient reformatting of the bibliography for different venues.
In terms of overall document formatting, the `rticles` package provides a suite of
custom R Markdown LaTeX formats suitable for many journals.

We have amended the Literate Programming section to mention these features briefly:

> Quarto and R Markdown have the ability to automatically generate a bibliography in many different
> journal styles [@noauthor_undated-pg].
> Further, dozens of journal-style Markdown document templates are freely available, and integrate
> with MyST notebook system or the `rticles` R package @Xie2018-rf.

@@

In light of this argument, we have added citations to Caetano and Aisenberg in the updated text.

### R3 Point 24

> 24. Page 11, line 39: It is not necessarily true that CSV files are better than Excel.
Excel can retain information about data types, for example, whereas CSVs do not.
It depends on what you are trying to accomplish.

RESPONSE: It is true that in some cases, Excel files are more convenient to view.
However, in the context of large data sets with thousands of rows and their analytical
reproducibility, Excel is a non-starter.
For example, it is well known that the behaviour of Excel changes between versions.
Earlier versions have limitations around the number of allowed rows and columns.
Small changes to Excel files such as merging cells can break machine readibility.
Excel users often use cell highlighting or text colour to encode some information which isn't
easily machine readable.
The simplicity of TSV and CSV formats means they are readily and rapidly ingested by commonly used
programming languages without needing to install extra packages.
In line with this and suggestions from others, the passage been amended:

> 5. Use file formats that are machine-readable and compatible with many different types of software.
> Some examples include comma- and tab-separated values (CSV/TSV) formats, eXtensible Markup Language (XML),
> JavaScript Object Notation (JSON), Hierarchical Data Format version 5 (HDF5) and Apache Parquet.

### 2023-08-31 manuscript/5pillars.Rmd
close #37 thanks Pierre

A number of guides recommending enhanced computational reproducibility practices have been developed
[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@Stodden2016-aj; @Beaulieu-Jones2017-vj; @Gruning2018-sg; @Akalin2018-eg;
@Brito2020-ld; @Peikert2021-eq; @Rodrigues2023-fw; @DuPre2022-bk; @Krafczyk2021-xv].
Inspired by these principles, we presents the *five pillars of reproducible computational research*
(**Figure 1**).
Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
transparent reporting and the role of computational environments.
The intended audience is anyone who analyzes biological data (not tool developers), however the
principles described here could equally apply to other domains of study.
There are clear parallels with the established three pillars of the open science framework,
(open data, code and papers) [@Gorgolewski2016-hv].

@@

Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
R, Python, Julia and other computer language chunks into documents [@kluyver2016; @granger2021; @Perkel2018-rz].
A noteworthy attempt to use literate programming in Jupyter notebooks to aim for reproducibility
has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
In this work, authors provided alongside the traditional research paper a
[Jupyter notebook](https://nbviewer.org/github/maayanlab/Zika-RNAseq-Pipeline/blob/master/Zika.ipynb)
that performed and documented the entire analysis procedure.
The MyST-NB (short for Markedly Structured Text notebook) is built on top of Jupyter notebooks or Markdown documents
and has some notable advantages around flexibility of content that make it an attractive choice for executable
article authoring [@DuPre2022-bk].

@@

* Flexible output formats.
These include PDF, DOC/DOCX and HTML, the latter having some notable benefits
including better usability for mobile devices like phones and tablets, and the ability to
support richer content such as interactive charts and dynamic tables (searchable, filterable, sortable).
Outputs include HTML documents, but also slideshows and even ebooks.
Quarto and R Markdown have the ability to automatically generate a bibliography in many different
journal styles [@noauthor_undated-pg].
Further, thanks to the `rticles` package, dozens of journal style R Markdown document templates are
freely available @Xie2018-rf.

@@

`git` is typically used at the command line, however it is also incorporated into integrated development
environments commonly used in bioinformatics including Rstudio, JupyterLab
[@Kluyver2016jupyter;@noauthor_undated-hg] and VS Code [@noauthor_undated-fn ;@Extensions_undated-ln].
There are also several `git` clients with graphical interfaces which better allow inspection
of code changes (e.g.: @noauthor_undated-rh, @noauthor_undated-kj).

@@

High throughput biology has already built a number of data repositories for specific data types mostly
conforming to FAIR principles.
Repositories like GEO, SRA, ENA and PRIDE were developed in the aftermath of the draft human
genome assembly and have become well known to omics researchers
[@Edgar2002-hs; @Kodama2012-bn; Martens2005-wo]).
The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
FAIR repositories that accepts data from other domains of study like ecology,
physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].
If no specialized repository exists, then a general-purpose repository such as
[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
should be considered. Researchers should be aware, however, that these repositories are often not moderated, 
and that it is their responsibility to make their data FAIR, in particular by providing sufficient metadata.
To this aim, general recommendations have been proposed to enhance data FAIRness in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].

@@

5. Use file formats that are machine-readable and compatible with many different types of software.
For example, the comma-separated values (CSV) and tab-separated values (TSV) formats are simple file
specifications that are suitable for very large data.

@@

Documentation is the glue that binds a data science project together.
The published article is the central artifact that outlines the research project and
links to the supporting materials.
From a reproducibility standpoint, the methods section is the most critical part.
It should be detailed enough so that other researchers can understand and replicate the
experiments/analysis and yield similar results.
Unfortunately, key details in bioinformatics data processing procedures are often omitted which limits
their reproducibility [@Nekrutenko2012; @Garijo2013-zl; @Simoneau2021-kp].
Commonly missing information includes versions of software and packages as well as any parameter setting
and configuration files.
The "Materials Design Analysis Reporting" (MDAR) checklist for authors has been developed to assist in
comprehensive methodological reporting in the life sciences [@Mellor2022-up], and "Minimum Information
About a Bioinformatics Investigation" guidelines describe good reporting practices for computational
biology research [@Tan2010-is].

@@

In addition, the `README` should outline the contents of the repository, how users can contribute to
the codebase and how to report issues, such as code errors and discrepancies.
The `README` should also list the contributors to the codebase and acknowledge any financial support.
It may also include a link to the research article, citation instructions and a mention of the software
license.
A recently described set of recommendations for making software FAIR describes further measures
that data analysts could use to enhance their code documentation @Patel2023-qr.

@@

When these principles are not integrated correctly and checked, problems can readily emerge
[@Barba2016-kk; @Samuel2022-ha; @Zaringhalam2020-ms].
This is why regular code testing after making updates to the code or data is considered best practice
[@Rule2019-eh; @The_Turing_Way_Community2019-ig; @Noor2022-uc].
For bioinformatics software, testing may involve a suite of unit tests of each function, and another
integration test whether the functions are working properly together [@Sarma2016-ol].

@@

In software development, continuous integration/continuous delivery (CI/CD) is a process of automated
software compilation and testing.
A derivation of this for data analysis has been termed "continuous analysis", and involves automatic
execution and testing whenever a change is made to a project's code base
[@Beaulieu-Jones2017-vj; @Blischak2019-dq].
We prefer the term "continuous validation" as it places more emphasis on the correctness of analyses
to ensure they are fit to share.
If a workflow is executed successfully and tests are passed, the repository and linked research article
can be decorated with a "badge" to inform users of the lack of errors in the code,
similar to badges used for other Open Science practices [@Kidwell2016-bb].
If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
article can be assessed computationally, then such tests could be used to demonstrate computational
reproducibility without the need for individuals to execute the workflow themselves [@Suetake2022-zq].

@@

Due to the relative lack of systematic reproducibility/replication studies in bioinformatics,
the severity of the reproducibility problem isn't well appreciated.
The studies that have been conducted point to reproducibility rates of ~10% in computational
life sciences [@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], but further studies
are required to bring attention to the issue.

@@

Individual researchers also face difficulties.
Life science graduates often receive inadequate training in data science that places them in a
disadvantage considering the data intensive nature of contemporary biology.
This is compounded by the strong demand for data science experts in industry and government which
leads to a deficit of academics fluent in data science
[@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
This makes it challenging for organisations to recruit and retain capable data analysts and cultivate
a community of practice.
Addressing this problem is not straightforward.
It requires ehancement of data science skills in undergraduate life science courses, investment in
upskilling the current workforce and offering working conditions that are competitive with other sectors.
Specific workshop-based training for current researchers could help [@Williams2023-ic].
The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
[@Wilson2014-mh].

@@

**Mark Ziemann** is a Senior Lecturer in Biotechnology and Bioinformatics at Deakin University.
He has recently been recruited to Burnet Institute as Head of Bioinformatics.
His research spans bioinformatics as applied to omics data analysis in biomedicine, as well as aspects
of research rigor.

>>>>

A number of guides recommending enhanced computational reproducibility practices have been developed
[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@Stodden2016-aj; @Wilson2017-dv; @Beaulieu-Jones2017-vj; @Gruning2018-sg; @Kim2018-nv; @Akalin2018-eg;
@Brito2020-ld; @Peikert2021-eq; @Rodrigues2023-fw; @DuPre2022-bk; @Krafczyk2021-xv].
Inspired by these principles, we presents the *five pillars of reproducible computational research*
(**Figure 1**).
Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
transparent reporting and the role of computational environments.
The intended audience is anyone who analyzes biological data (not tool developers), however the
principles described here could equally apply to other domains of study.
There are clear parallels with the established three pillars of the open science framework,
(open data, code and papers) [@Gorgolewski2016-hv].

@@

Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
R, Python, Julia and other computer language chunks into documents [@kluyver2016; @granger2021; @Perkel2018-rz].
A noteworthy attempt to use literate programming in Jupyter notebooks to aim for reproducibility
has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
In this work, authors provided alongside the traditional research paper a
[Jupyter notebook](https://nbviewer.org/github/maayanlab/Zika-RNAseq-Pipeline/blob/master/Zika.ipynb)
that performed and documented the entire analysis procedure.
The MyST (short for Markedly Structured Text notebook) is built on top of Jupyter notebooks or Markdown documents
and has some notable advantages around flexibility of content that make it an attractive choice for executable
article authoring [@DuPre2022-bk].

@@

* Flexible output formats.
These include PDF, DOC/DOCX and HTML, the latter having some notable benefits
including better usability for mobile devices like phones and tablets, and the ability to
support richer content such as interactive charts and dynamic tables (searchable, filterable, sortable).
Outputs include HTML documents, but also slideshows and even ebooks.
Quarto and R Markdown have the ability to automatically generate a bibliography in many different
journal styles [@noauthor_undated-pg].
Further, dozens of journal-style Markdown document templates are freely available, and integrate
with MyST notebook system or the `rticles` R package @Xie2018-rf.

@@

`git` is typically used at the command line, however it is also incorporated into integrated development
environments commonly used in bioinformatics including Rstudio, JupyterLab
[@Kluyver2016jupyter;@noauthor_undated-hg] and VS Code [@noauthor_undated-fn ;@Extensions_undated-ln].
There are also several `git` clients with graphical interfaces which better allow inspection
of code changes (e.g.: [@noauthor_undated-rh; @noauthor_undated-kj]).

@@

High throughput biology has already built a number of data repositories for specific data types mostly
conforming to FAIR principles.
Repositories like GEO, SRA, ENA and PRIDE were developed in the aftermath of the draft human
genome assembly and have become well known to omics researchers
[@Edgar2002-hs; @Kodama2012-bn; @Martens2005-wo].
The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
FAIR repositories that accepts data from other domains of study like ecology,
physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].
If no specialized repository exists, then a general-purpose repository such as
[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
should be considered. Researchers should be aware, however, that these repositories are often not moderated, 
and that it is their responsibility to make their data FAIR, in particular by providing sufficient metadata.
To this aim, general recommendations have been proposed to enhance data FAIRness in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].

@@

5. Use file formats that are machine-readable and compatible with many different types of software.
Some examples include comma- and tab-separated values (CSV/TSV) formats, eXtensible Markup Language (XML),
JavaScript Object Notation (JSON), Hierarchical Data Format version 5 (HDF5) and Apache Parquet.

@@

Documentation is the glue that binds a data science project together.
The published article is the central artifact that outlines the research project and
links to the supporting materials.
From a reproducibility standpoint, the methods section is the most critical part.
It should be detailed enough so that other researchers can understand and replicate the
experiments/analysis and yield similar results.
Unfortunately, key details in bioinformatics data processing procedures are often omitted which limits
their reproducibility [@Nekrutenko2012; @Garijo2013-zl; @Kim2018-nv; @Simoneau2021-kp].
Commonly missing information includes versions of software and packages as well as any parameter setting
and configuration files.
The "Materials Design Analysis Reporting" (MDAR) checklist for authors has been developed to assist in
comprehensive methodological reporting in the life sciences [@Mellor2022-up], and "Minimum Information
About a Bioinformatics Investigation" guidelines describe good reporting practices for computational
biology research [@Tan2010-is].

@@

In addition, the `README` should outline the contents of the repository, how users can contribute to
the project and how to report issues, such as code errors and discrepancies.
The `README` should also list the contributors to the codebase and acknowledge any financial support.
It may also include a link to the research article, citation instructions and a mention of the software
license.
A recently described set of recommendations for making software FAIR describes further measures
that data analysts could use to enhance their code documentation @Patel2023-qr.

@@

When these principles are not integrated correctly and checked, problems can readily emerge
[@Barba2016-kk; @Samuel2022-ha; @Zaringhalam2020-ms].
This is why regular code testing after making updates to the code or data is considered best practice
[@Kim2018-nv; @Rule2019-eh; @The_Turing_Way_Community2019-ig; @Noor2022-uc].
For bioinformatics software, testing may involve a suite of unit tests of each function, and another
integration test whether the functions are working properly together [@Sarma2016-ol].

@@

In software development, continuous integration/continuous delivery (CI/CD) is a process of automated
software compilation and testing.
A derivation of this for data analysis has been termed "continuous analysis", and involves automatic
execution and testing whenever a change is made to a project's code base
[@Beaulieu-Jones2017-vj; @Blischak2019-dq].
We prefer the term "continuous validation" as it places more emphasis on the correctness of analyses
to ensure they are fit to share.
If a workflow is executed successfully and tests are passed, the repository and linked research article
can be decorated with a "badge" to indicate computational reproducibility,
similar to badges used for other Open Science practices [@Kidwell2016-bb].
If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
article can be assessed computationally, then such tests could be used to demonstrate computational
reproducibility without the need for individuals to execute the workflow themselves [@Suetake2022-zq].

@@

Due to the relative lack of systematic reproducibility/replication studies in bioinformatics,
the severity of the reproducibility problem isn't well appreciated.
The studies that have been conducted point to reproducibility rates of ~20% in computational
life sciences [@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], but further studies
are required to bring attention to the issue.

@@

Individual researchers also face difficulties.
Life science graduates often receive inadequate training in data science that places them in a
disadvantage considering the data intensive nature of contemporary biology.
This is compounded by the strong demand for data science experts in industry and government which
leads to a deficit of academics fluent in data science
[@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
This makes it challenging for organisations to recruit and retain capable data analysts and cultivate
a community of practice.
Addressing this problem is not straightforward.
It requires ehancement of reproducible data science skills in undergraduate life science courses,
investment in upskilling the current workforce and offering working conditions that are competitive
with other sectors.
Specific workshop-based training for current researchers could help [@Williams2023-ic].
The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
[@Wilson2014-mh].

@@

**Mark Ziemann** is Head of Bioinformatics at Burnet Institute and former Lecturer in
Biotechnology and Bioinformatics at Deakin University.
His research spans bioinformatics as applied to omics data analysis in biomedicine, as well as aspects
of research rigor.
